Effective multi-body SU($N$)-symmetric interactions of ultracold fermionic atoms on a 3-D lattice

### 2019-02-02 multi_body_interactions/papers/theory/paper.tex
updated multi-body paper with edits in response to referee comments

@@

@@

\begin{abstract}
  Rapid advancements in the experimental capabilities with ultracold
  alkaline-earth-like atoms (AEAs) bring to a surprisingly near term
  the prospect of performing quantum simulations of spin models and
  lattice field theories exhibiting SU($N$) symmetry.  Motivated in
  particular by recent experiments preparing high density samples of
  strongly interacting ${}^{87}$Sr atoms in a three-dimensional
  optical lattice, we develop a low-energy effective theory of
  fermionic AEAs which exhibits emergent multi-body SU($N$)-symmetric
  interactions, where $N$ is the number of atomic nuclear spin levels.
  We fully characterize the low-lying excitations in our effective
  theory, and compare predictions of many-body interaction energies
  with direct measurements of many-body excitation spectra in an
  optical lattice clock.  Our work makes the first step in enabling a
  controlled, bottom-up experimental investigation of multi-body
  SU($N$) physics.
\end{abstract}

@@

\tableofcontents

@@

% Rapid advancements in the experimental capabilities with ultracold
% alkaline-earth-like atoms (AEAs) \cite{fukuhara2007degenerate,
% stellmer2009boseeinstein, deescobar2009boseeinstein,
% tey2010doubledegenerate, desalvo2010degenerate, taie2012su,
% martin2013quantum, cappellini2014direct, scazza2014observation,
% zhang2014spectroscopic, cazalilla2014ultracold} bring to a
% surprisingly near term the prospect of performing quantum
% simulations of spin models and lattice field theories exhibiting
% SU($N$) symmetry \cite{cazalilla2009ultracold, hermele2009mott,
% hermele2011topological, gorshkov2010twoorbital, banerjee2013atomic,
% cazalilla2014ultracold}.  AEAs have ground and metastable
% (electronic) orbital states\note{[I want to briefly discuss how we
% present this]} ${}^1S_0$ and ${}^3P_0$, often referred to as the
% ``clock'' states due to their use in optical lattice clocks
% \cite{derevianko2011colloquium, katori2011optical}, with
% ultra-narrow (millihertz) optical transition linewidths and
% associated $\sim10^2$ second excited-state lifetimes.  Due to the
% lack of electronic orbital angular momentum in the ${}^1S_0$ and
% ${}^3P_0$ states, AEAs exhibit decoupled orbital and nuclear spin
% degrees of freedom.  For atoms with total nuclear spin $I$ and
% $N=2I+1$ nuclear spin levels, this decoupling leads to
% nuclear-spin-conserving SU($N$)-symmetric interactions governed
% entirely by orbital-state parameters \cite{cazalilla2009ultracold,
% taie2012su, zhang2014spectroscopic}.

@@

In 2015 the ${}^{87}$Sr optical lattice clock (OLC) at JILA, operated
in a one-dimensional (1-D) lattice at microkelvin temperatures,
achieved a total fractional uncertainty of $2\times10^{-18}$
\cite{bloom2014optical, nicholson2015systematic}.  Last year (2017), a
new generation of OLCs became operational at JILA, interrogating a
Fermi degenerate gas of ${}^{87}$Sr atoms in a 3-D lattice at
nanokelvin temperatures \cite{campbell2017fermidegenerate}.  All of
these atoms' degrees of freedom, including the electronic orbital,
nuclear spin, and motional states, can be fully controlled with high
fidelity in a 3-D lattice \cite{daley2008quantum,
  gorshkov2009alkalineearthmetal, daley2011quantum,
  daley2011statedependent}.  With frequency measurements reaching the
$10^{-19}$ fractional uncertainty level, the new OLCs are thus
enabling an exciting opportunity to probe, for the first time, quantum
dynamics with sub-millihertz spectral resolution
\cite{campbell2017fermidegenerate}.

@@

% One of the major drivers of experimental capabilities with AEAs is
% the development of optical lattice clocks
% \cite{derevianko2011colloquium, katori2011optical}.  These clocks
% traditionally operate in a one-dimensional (1-D) geometry with
% harmonic confinement of atoms in the plane orthogonal to the lattice
% axis \cite{derevianko2011colloquium, hinkley2013atomic,
% bloom2014optical}.  By operating at higher atomic densities, one can
% simultaneously interrogate a larger number of atoms and thereby
% increase the signal-to-noise ratio of the clock.  With the extreme
% sensitivities accessible to state-of-the-art optical lattice clocks,
% however, 1-D clock experiments with ${}^{87}$Sr and ${}^{171}$Yb
% detect density-dependent fractional frequency shifts at the
% $10^{-16}$ level \cite{campbell2009probing, lemke2009spin}.  For the
% purpose of timekeeping, these frequency shifts amount to an
% undesirable source of systematic error, but they introduced a new,
% exciting paradigm for the field of quantum simulation: a marriage
% between precision clock spectroscopy and quantum many-body physics
% \cite{taie2012su, scazza2014observation, martin2013quantum,
% cappellini2014direct, zhang2014spectroscopic, rey2014probing}.
% Somewhat counter-intuitively, one of the most promising strategies
% for improving the interaction-limited 1-D clock design has been to
% use higher dimensional lattices to generate tighter on-site
% confinement of atoms, thereby {\it increasing} the strength of
% inter-atomic interactions and making evolution into interacting
% many-body states energetically unfavorable
% \cite{swallows2011suppression, lemke2011wave,
% campbell2017fermidegenerate}.  A wonderful side-effect of this
% strategy for building better clocks is the development of a highly
% controllable quantum many-body system in the strongly-interacting
% regime, where inter-atomic interactions set the largest energy scale
% relevant for system dynamics.

In this work, we investigate the first experimental capabilities with
ultracold fermionic AEAs to prepare high-density samples in a 3-D
lattice with multiple occupation of individual lattice sites
\cite{goban2018emergence}.  Specifically, we consider the preparation
of isolated few-body systems in the deep-lattice limit, and carry out
a bottom-up investigation of emergent multi-body interactions on
multiply-occupied lattice sites.  These multi-body interactions appear
in a low-energy effective theory of the atoms, and inherit the SU($N$)
symmetry of their bare, pair-wise interactions, thereby enabling
experimental studies of multi-body SU($N$) physics through the
exquisite capabilities with OLCs.

@@

In the absence of hyperfine coupling, as when addressing the spinless
${}^1S_0$ ($\g$) and ${}^3P_0$ ($\e$) orbital states, interactions
between any two atoms are governed by their orbital states alone, and
are therefore characterized by four scattering lengths $a_X$ with
$X\in\set{\g\g,\e\g^-,\e\g^+,\e\e}$, where the $+$ ($-$) superscript
denotes symmetrization (anti-symmetrization) of a two-body orbital
state under particle exchange.  In the low-energy limit, we can write
the bare two-body interaction Hamiltonian in the form
\cite{gorshkov2010twoorbital}
\begin{align}
  H_{\t{int}} = \sum_{\substack{\mu<\nu\\s}}
  G_s \int d^3x~ \hat\rho_{\mu s} \hat\rho_{\nu s}
  + G_+ \sum_{\mu,\nu}
  \int d^3x~ \hat\rho_{\mu,\e} \hat\rho_{\nu,\g}
  + G_- \sum_{\mu,\nu} \int d^3x~
  \hat\psi_{\mu,\e}^\dag \hat\psi_{\nu,\g}^\dag
  \hat\psi_{\nu,\e} \hat\psi_{\mu,\g},
\end{align}
where $\hat\psi_{\mu s}$ is a fermionic field operator for atoms with
nuclear spin $\mu$ and orbital state $s$;
$\hat\rho_{\mu s}\equiv\hat\psi_{\mu s}^\dag\hat\psi_{\mu s}$ is an
atomic density field operator; and the coupling constants $G_X$ are
defined in terms of the scattering lengths by
\begin{align}
  G_{s=\g,\e} \equiv \f{4\pi}{m_A}~ a_{ss}, &&
  G_\pm \equiv \f{2\pi}{m_A}~ \p{a_{\e\g+} \pm a_{\e\g-}},
  \label{eq:couplings}
\end{align}
where $m_A$ is the mass of a single atom.  Defining for brevity
\begin{align}
  G^{qr}_{st} \equiv \left\{
    \begin{array}{ll}
      G_q & ~ q = r = s = t \\
      G_+ & ~ q \ne r ~ \t{and} ~ (q,r) = (s,t) \\
      G_- & ~ q \ne r ~ \t{and} ~ (q,r) = (t,s) \\
      0 & ~ \t{otherwise}
    \end{array}\right.,
  \label{eq:coupling_tensor}
\end{align}
where $q,r,s,t\in\set{\g,\e}$ are orbital state indices, we can
alternately write the bare two-body interaction Hamiltonian in the
more compact form
\begin{align}
  H_{\t{int}}
  = \f12 \sum_{\substack{q,r,s,t\\\mu,\nu}} G^{qr}_{st} \int d^3x~
  \hat\psi_{\mu s}^\dag \hat\psi_{\nu t}^\dag
  \hat\psi_{\nu r} \hat\psi_{\mu q}.
  \label{eq:H_int_fields}
\end{align}
For nuclear spins $\mu,\nu$, the symbol $G^{qr}_{st}$ gives the
coupling constant between the two-atom states
$(\mu,q)+(\nu,r)\leftrightarrow(\mu,s)+(\nu,t)$.

We now expand the field operators $\hat\psi_{\mu s}$ in the Wannier
basis for a 3-D lattice, such that
$\hat\psi_{\mu s}(x) = \sum_{i,n} \phi_{in}(x) \c_{in\mu s}$ with
spatial wavefunctions $\phi_{in}$ and fermionic annihilation operators
$\c_{in\mu s}$ indexed by lattice sites $i$ and motional states $n$.
Invoking the tight-binding approximation, we assume that the spatial
overlap integral in \eqref{eq:H_int_fields} is negligible unless all
wavefunctions are localized at the same lattice site; we discuss the
breakdown of this approximation and its consequences for our
low-energy effective theory in Appendix \ref{sec:error}.  The relevant
spatial overlap integral is then
\begin{align}
  K^{k\ell}_{mn}
  \equiv \int d^3x~ \phi_{im}^* \phi_{in}^* \phi_{i\ell} \phi_{ik},
  \label{eq:K_klmn}
\end{align}
which for a lattice with discrete translational invariance is
independent of the lattice site $i$.  The two-body interaction
Hamiltonian can be written in terms of this overlap integral as
\begin{align}
  H_{\t{int}}
  = \f12 \sum_{\substack{i,k,\ell,m,n\\q,r,s,t\\\mu,\nu}}
  K^{k\ell}_{mn} G^{qr}_{st}
  \c_{im\mu s}^\dag \c_{in\nu t}^\dag \c_{i\ell\nu r} \c_{ik\mu q}
  \equiv \f12 \sum K^{k\ell}_{mn} G^{qr}_{st}
  \c_{m\mu s}^\dag \c_{n\nu t}^\dag \c_{\ell\nu r} \c_{k\mu q},
  \label{eq:H_int}
\end{align}
where for brevity we will henceforth suppress the identical site index
($i$) on all operators, and implicitly sum over all free indices in a
summand (i.e.~indices which do not have a fixed value).  We may also
at times suppress motional state indices on the overlap integral
$K^{k\ell}_{mn}$, in which case the suppressed indices are implicitly
zero (corresponding to a motional ground state); i.e.
\begin{align}
  K^{\ell m}_n \equiv K^{\ell m}_{n,0},
  &&
  K^m_n \equiv K^{m,0}_{n,0},
  &&
  K_{mn} \equiv K^{0,0}_{mn},
  &&
  K_n \equiv K^{0,0}_{n,0},
  &&
  K \equiv K^{0,0}_{0,0}.
  \label{eq:K}
\end{align}
For simplicity, we will also generally work in a gauge for which all
two-body overlap integrals are real, such that
$K^{k\ell}_{mn}={K^{k\ell}_{mn}}^*=K^{mn}_{k\ell}$.  The existence of
such a gauge is guaranteed by the analytic properties of the Wannier
orbitals $\phi_{in}$ \cite{kohn1959analytic}.

We are interested in the ground and low-lying orbital eigenstates of
several interacting atoms on a single lattice site, as such states can
be readily prepared and coherently addressed in current experiments
with ultracold atoms \cite{goban2018emergence}.  Although these
experiments operate at temperatures sufficiently low for negligible
thermal occupation of excited motional states, simply neglecting
excited single-particle motional states fails to reproduce the
observed spectra of multiply-occupied lattice sites.  The reason for
this failure lies in the fact that the bare two-body Hamiltonian
$H_{\t{int}}$ in \eqref{eq:H_int} is not diagonal in the basis of
single-particle motional states.  Even at zero temperature, therefore,
the presence of excited motional states modifies many-body atomic
energy spectra when the magnitude of off-diagonal elements of
$H_{\t{int}}$ approach the scale of low-lying single-particle motional
state energy gaps.  To address this problem, we develop a low-energy
effective theory for ultracold two-level fermionic atoms on a lattice.
As our work is motivated by progress in experiments which initialize
all atoms in identical (i.e.~ground) orbital and motional states, to
simplify our theory we assume that any $N$ atoms on a single lattice
site occupy $N$ nuclear spin states.  Multiple occupation of a single
nuclear spin state on the same lattice site is initially forbidden by
fermionic statistics, and subsequently cannot be violated in the
absence of tunneling and hyperfine coupling.

Our low-energy effective theory exhibits SU($N$)-symmetric multi-body
interactions, such that the effective interaction Hamiltonian can be
written in the form
\begin{align}
  H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} H_M,
\end{align}
where $H_M$ is an $M$-body Hamiltonian, $I$ is the total nuclear spin
of each atom (e.g.~$I=9/2$ for ${}^{87}$Sr), and the sum terminates at
$2I+1$ because this is the largest number of atoms which may initially
occupy a single lattice site.  Focusing on the low-lying orbital state
excitations of this theory, we find that SU($N$) symmetry allows us to
express multi-body Hamiltonians in the simple form
\begin{align}
  H_M = \sum_{\abs{\set{\mu_j}}=M}
  \p{U_{M,\g} \prod_{\alpha=1}^M \n_{\mu_\alpha,\g}
    + U_{M,+} \n_{\mu_1,\e} \prod_{\alpha=2}^M \n_{\mu_\alpha,\g}
    + U_{M,-} \c_{\mu_1,\g}^\dag \c_{\mu_2,\e}^\dag
    \c_{\mu_2,\g} \c_{\mu_1,\e} \prod_{\alpha=3}^M \n_{\mu_\alpha,\g}},
  \label{eq:H_M_preview}
\end{align}
where $\c_{\mu s}\equiv\c_{0,\mu s}$ is a fermionic annihilation
operator for an atom in the motional ground state with nuclear spin
$\mu$ and orbital state $s$;
$\n_{\mu s}\equiv \c_{\mu s}^\dag \c_{\mu s}$ is a number operator;
and the sum is performed over all choices of nuclear spins $\mu_j$
with $j=1,2,\cdots,M$ for which all $\mu_j$ are distinct, or
equivalently all choices of $\mu_j$ for which the set $\set{\mu_j}$
contains $M$ elements, for a total of ${2I+1\choose M}\times\p{M!}$
nuclear spin combinations.  The coefficients $U_X$ can be expressed in
terms of the coupling constants of the bare two-body interaction
Hamiltonian (see Appendix \ref{sec:U_X}).  The key feature of the
$M$-body interactions in \eqref{eq:H_M_preview} is that they take a
form identical to the two-body interactions between motional
ground-state atoms, but with the addition of $M-2$ spectator atoms.
This form is a direct consequence of the SU($N$) symmetry of
underlying two-body interactions.

@@

As the bare two-body Hamiltonian $H_{\t{int}}$ is not diagonal with
respect to single-particle motional states, the problem of determining
many-body spectra and eigenstates nominally involves all motional
degrees of freedom of the atoms involved.  If we are interested in
addressing internal atomic states at zero temperature, however, in
principle we need only to consider the motional ground state
associated with each many-body internal atomic state.  We can then
ignore all excited many-body motional states, as they will be neither
thermally occupied nor externally interrogated.  This procedure makes
the many-body problem more tractable by greatly reducing the
dimensionality of our the relevant Hilbert space.

@@

two-body interaction.  In general, the sums over excited states in the
loop diagrams of \eqref{eq:H_2_expansion} will diverge
\cite{johnson2012effective}, which implies the need to renormalize the
coupling constants in our theory.  Even if these diagrams do not
diverge, we would like to express the effective two-body interactions
in terms of net, physically observable two-body coupling constants as
on the right-hand side of \eqref{eq:H_2_expansion}, rather than in
terms of long sums at all orders of the bare coupling constants.  We
therefore renormalize our coupling constants by introducing

@@

via $G_X\to G_X+\tilde G_X$.  The values of these counter-terms are
fixed by enforcing that $G_X$ is equal to the net two-body interaction
strength.  Diagrammatically, this renormalization condition is

@@

a vertex. To leading order in the coupling constants, it follows that
the counter-terms are second order in the coupling constants,
i.e.~$\tilde G\sim G^2$.  This finding implies that counter-terms only
come into play at next-to-leading order (NLO) in the calculation of
any effective $M$-body interaction Hamiltonian.  By construction,
these counter-terms exactly cancel all terms beyond leading order in

@@

Second, the overview we provided in section \ref{sec:overview}, and
consequently our result in \eqref{eq:H_2}, neglects the effect of
momentum-dependent two-body scattering.  When the effective range of
momentum-dependent interactions is comparable to the scattering
lengths $a_X$, as is the case for ultracold ${}^{87}$Sr, these
interactions are third order in the coupling constants $G_X$.  Just as
the $\O\p{G^2}$ counter-terms do not affect $M$-body interactions
until next-to-leading order (NLO), the $\O\p{G^3}$ momentum-dependent
interactions do not come into play until next-next-leading order
(NNLO).  Given that we develop our low-energy effective theory through
third order in the coupling constants, these interactions will not
appear in any of our three- and four-body calculations, but they do
need to be considered in the calculation of pair-wise interaction
energies.  The primary interest of our work, however, concerns
effective $M$-body interactions for $M\ge3$; we therefore defer the
calculation of momentum-dependent two-body interactions to Appendix
\ref{sec:momentum_dependence}.

@@

>>>>

\newcommand{\mA}{m_{\text{A}}} % symbol for the mass of an atom

@@

\usepackage{setspace} % to modify spacing in table of contents

@@

\begin{abstract}
  Rapid advancements in the experimental capabilities with ultracold
  alkaline-earth-like atoms (AEAs) bring to a surprisingly near term
  the prospect of performing quantum simulations of spin models and
  lattice field theories exhibiting SU($N$) symmetry.  In order to
  analyze recent experiments preparing high density samples of
  strongly interacting ${}^{87}$Sr atoms in a three-dimensional
  optical lattice, we develop a low-energy effective theory of
  fermionic AEAs which exhibits emergent multi-body SU($N$)-symmetric
  interactions, where $N$ is the number of atomic nuclear spin levels.
  We fully characterize the low-lying excitations in our effective
  theory, and compare predictions of many-body interaction energies
  with direct measurements of many-body excitation spectra in an
  optical lattice clock.  Our work makes the first step in enabling a
  controlled, bottom-up experimental investigation of multi-body
  SU($N$) physics.
\end{abstract}

@@

\begin{spacing}{1.3}
  \tableofcontents
\end{spacing}

@@

@@

In 2015 the ${}^{87}$Sr optical lattice clock (OLC) at JILA, operated
in a one-dimensional (1-D) lattice at microkelvin temperatures,
achieved a total fractional uncertainty of $2\times10^{-18}$
\cite{bloom2014optical, nicholson2015systematic}.  More recently
(2017), a new generation of OLCs became operational at JILA,
interrogating a Fermi degenerate gas of ${}^{87}$Sr atoms in a 3-D
lattice at nanokelvin temperatures \cite{campbell2017fermidegenerate}.
All of these atoms' degrees of freedom, including the electronic
orbital, nuclear spin, and motional states, can be fully controlled
with high fidelity in a 3-D lattice \cite{daley2008quantum,
  gorshkov2009alkalineearthmetal, daley2011quantum,
  daley2011statedependent}.  With frequency measurements reaching the
$10^{-19}$ fractional uncertainty level, the new OLCs are thus
enabling an exciting opportunity to probe, for the first time, quantum
dynamics with sub-millihertz spectral resolution
\cite{campbell2017fermidegenerate}.

@@

In this work, we investigate the first experimental capabilities with
ultracold fermionic AEAs to prepare high-density samples in a 3-D
lattice with multiple occupation of individual lattice sites
\cite{goban2018emergence}.  Specifically, we consider ground-state
preparation of isolated few-body systems in the deep-lattice limit,
and carry out a bottom-up investigation of emergent multi-body
interactions on multiply-occupied lattice sites.  These multi-body
interactions appear in a low-energy effective theory of the atoms, and
inherit the SU($N$) symmetry of their bare, pair-wise interactions,
thereby enabling experimental studies of multi-body SU($N$) physics
through the exquisite capabilities with OLCs.  We briefly summarize
the experimental procedures relevant to our work in Appendix
\ref{sec:experiment_summary}.

@@

In the absence of hyperfine coupling, as when addressing the spinless
${}^1S_0$ ($\g$) and ${}^3P_0$ ($\e$) orbital states of AEAs,
interactions between any two atoms are governed by their orbital
states alone, and are therefore characterized by four scattering
lengths $a_X$ with $X\in\set{\g\g,\e\g^-,\e\g^+,\e\e}$, where the $+$
($-$) superscript denotes symmetrization (anti-symmetrization) of a
two-body orbital state under particle exchange.  In the low-energy
limit, we can write the bare two-body interaction Hamiltonian in the
form \cite{gorshkov2010twoorbital}
\begin{align}
  H_{\t{int}} = \sum_{\substack{\mu<\nu\\s}}
  G_s \int d^3x~ \hat\rho_{\mu s} \hat\rho_{\nu s}
  + G_+ \sum_{\mu,\nu}
  \int d^3x~ \hat\rho_{\mu,\e} \hat\rho_{\nu,\g}
  + G_- \sum_{\mu,\nu} \int d^3x~
  \hat\psi_{\mu,\e}^\dag \hat\psi_{\nu,\g}^\dag
  \hat\psi_{\nu,\e} \hat\psi_{\mu,\g},
  \label{eq:H_int_start}
\end{align}
where $\hat\psi_{\mu s}$ is a fermionic field operator for atoms with
nuclear spin $\mu$ and orbital state $s$;
$\hat\rho_{\mu s}\equiv\hat\psi_{\mu s}^\dag\hat\psi_{\mu s}$ is an
atomic density field operator; and the coupling constants $G_X$ are
defined in terms of the scattering lengths by
\begin{align}
  G_{s=\g,\e} \equiv \f{4\pi}{\mA}~ a_{ss}, &&
  G_\pm \equiv \f{2\pi}{\mA}~ \p{a_{\e\g+} \pm a_{\e\g-}},
  \label{eq:couplings}
\end{align}
where $\mA$ is the mass of a single atom.  Defining for brevity
\begin{align}
  G^{qr}_{st} \equiv \left\{
    \begin{array}{ll}
      G_q & ~ q = r = s = t \\
      G_+ & ~ q \ne r ~ \t{and} ~ (q,r) = (s,t) \\
      G_- & ~ q \ne r ~ \t{and} ~ (q,r) = (t,s) \\
      0 & ~ \t{otherwise}
    \end{array}\right.,
  \label{eq:coupling_tensor}
\end{align}
where $q,r,s,t\in\set{\g,\e}$ are orbital state indices, we can
alternately write the bare two-body interaction Hamiltonian in the
more compact form
\begin{align}
  H_{\t{int}}
  = \f12 \sum_{\substack{q,r,s,t\\\mu,\nu}} G^{qr}_{st} \int d^3x~
  \hat\psi_{\mu s}^\dag \hat\psi_{\nu t}^\dag
  \hat\psi_{\nu r} \hat\psi_{\mu q}.
  \label{eq:H_int_fields}
\end{align}
For nuclear spins $\mu,\nu$, the symbol $G^{qr}_{st}$ gives the
coupling constant between the two-atom states
$(\mu,q)+(\nu,r)\leftrightarrow(\mu,s)+(\nu,t)$.  Note that
\eqref{eq:H_int_start} and \eqref{eq:H_int_fields} neglect
momentum-dependent corrections to $s$-wave interactions, which are
generally captured (in a low-energy limit) by the use of an
energy-dependent interaction pseudo-potential \cite{blume2002fermi}.
These corrections will appear only at third order in our effective
theory, as corrections to two-body interaction energies, and do not
affect our discussions of multi-body interactions.  We therefore defer
further discussion of momentum-dependent corrections to the end of
section \ref{sec:two_body}, and their calculation to Appendix
\ref{sec:momentum_dependence}.

We now expand the field operators $\hat\psi_{\mu s}$ in the Wannier
basis for a 3-D lattice, such that
$\hat\psi_{\mu s}(x) = \sum_{i,n} \phi_{in}(x) \c_{in\mu s}$ with
spatial wavefunctions $\phi_{in}$ and fermionic annihilation operators
$\c_{in\mu s}$ indexed by lattice sites $i$ and motional states $n$.
Invoking the tight-binding approximation, we assume that the spatial
overlap integral in \eqref{eq:H_int_fields} is negligible unless all
wavefunctions are localized at the same lattice site; we discuss the
breakdown of this approximation and its consequences for our
low-energy effective theory in Appendix \ref{sec:error}.  The relevant
spatial overlap integral is then
\begin{align}
  K^{k\ell}_{mn}
  \equiv \int d^3x~ \phi_{im}^* \phi_{in}^* \phi_{i\ell} \phi_{ik},
  \label{eq:K_klmn}
\end{align}
which for a lattice with discrete translational invariance is
independent of the lattice site $i$.  The two-body interaction
Hamiltonian can be written in terms of this overlap integral as
\begin{align}
  H_{\t{int}}
  = \f12 \sum_{\substack{i,k,\ell,m,n\\q,r,s,t\\\mu,\nu}}
  K^{k\ell}_{mn} G^{qr}_{st}
  \c_{im\mu s}^\dag \c_{in\nu t}^\dag \c_{i\ell\nu r} \c_{ik\mu q}
  \equiv \f12 \sum K^{k\ell}_{mn} G^{qr}_{st}
  \c_{m\mu s}^\dag \c_{n\nu t}^\dag \c_{\ell\nu r} \c_{k\mu q},
  \label{eq:H_int}
\end{align}
where for brevity we will henceforth suppress the identical site index
($i$) on all operators, and implicitly sum over all free indices in a
summand (i.e.~indices which do not have a fixed value).  We may also
at times suppress motional state indices on the overlap integral
$K^{k\ell}_{mn}$, in which case the suppressed indices are implicitly
zero (corresponding to a motional ground state); i.e.
\begin{align}
  K^{\ell m}_n \equiv K^{\ell m}_{n,0},
  &&
  K^m_n \equiv K^{m,0}_{n,0},
  &&
  K_{mn} \equiv K^{0,0}_{mn},
  &&
  K_n \equiv K^{0,0}_{n,0},
  &&
  K \equiv K^{0,0}_{0,0}.
  \label{eq:K}
\end{align}
For simplicity, we will also generally work in a gauge for which all
two-body overlap integrals are real, such that
$K^{k\ell}_{mn}={K^{k\ell}_{mn}}^*=K^{mn}_{k\ell}$.  The existence of
such a gauge is guaranteed by the analytic properties of the Wannier
wavefunctions $\phi_{in}$ \cite{kohn1959analytic}.

Current experiments with ${}^{87}$Sr can prepare up to five atoms
atoms in the same (ground) orbital state on a single lattice site, and
coherently address states with a single orbital excitation per lattice
site (see Appendix \ref{sec:experiment_summary})
\cite{goban2018emergence}.  The temperature of the atoms in these
experiments is much smaller than single-atom motional excitation
energies $\Delta_n\equiv E_n-E_0$ for $n>0$, which implies a
negligible thermal occupation of excited motional states.  A naive
calculation of the multi-body orbital excitation spectrum, however,
assuming that atoms only occupy motional ground states with spatial
wavefunctions $\phi_{i,0}$, fails to reproduce the experimentally
observed orbital excitation spectrum of multiply-occupied lattice
sites.  The reason that such a calculation fails to reproduce the
observed excitation spectrum has to do with the fact that the two-body
Hamiltonian $H_{\t{int}}$ in \eqref{eq:H_int} is not diagonal in the
basis of single-particle motional eigenstates (i.e.~$\phi_{in}$).  For
interacting atoms, the true motional ground states are therefore
admixtures of the non-interacting motional eigenstates; when the
off-diagonal elements of $H_{\t{int}}$ are comparable in magnitude to
the excitation energies $\Delta_n$, this admixture can have
non-negligible weight outside the non-interacting motional
ground-state subspace.  In order to compute the correct many-body
orbital excitation spectrum, we develop a low-energy effective theory
for such atoms.  As our work is motivated by experiments that
initialize all atoms in identical (i.e.~ground) orbital and motional
states, to simplify our theory we assume that any $N$ atoms on a
single lattice site occupy $N$ nuclear spin states.  Ground-state
preparation implies that multiple occupation of a single nuclear spin
state on any given lattice site is initially forbidden by fermionic
statistics, and subsequently cannot occur in the absence of inter-site
effects or hyperfine coupling.

Our low-energy effective theory exhibits SU($N$)-symmetric multi-body
interactions, such that the effective interaction Hamiltonian can be
written in the form
\begin{align}
  H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} \sum_{p\ge1} H_M^{(p)},
\end{align}
where $H_M^{(p)}$ is an $M$-body Hamiltonian of order $p$ in the
coupling constants $G^{qr}_{st}$, $I$ is the total nuclear spin of
each atom (e.g.~$I=9/2$ for ${}^{87}$Sr), and the sum terminates at
$2I+1$ because this is the largest number of atoms which may initially
occupy a single lattice site.  We explicitly compute all $M$-body
Hamiltonians $H_M\equiv\sum_p H_M^{(p)}$ through order $p=3$, yielding
effective two-, three-, and four-body interactions.  To facilitate
understanding and comparison with experimental measurements of
density-dependent atomic energy spectra [34], in which states with
multiple orbital excitations on a single lattice site are
spectroscopically off-resonant, we also consider a restriction of our
theory to states with at most one orbital excitation per lattice site.
In this case, we find that SU($N$) symmetry allows us to express
multi-body Hamiltonians in the simple form
\begin{align}
  H_M = \sum_{\abs{\set{\mu_j}}=M}
  \p{U_{M,\g} \prod_{\alpha=1}^M \n_{\mu_\alpha,\g}
    + U_{M,+} \n_{\mu_1,\e} \prod_{\alpha=2}^M \n_{\mu_\alpha,\g}
    + U_{M,-} \c_{\mu_1,\g}^\dag \c_{\mu_2,\e}^\dag
    \c_{\mu_2,\g} \c_{\mu_1,\e} \prod_{\alpha=3}^M \n_{\mu_\alpha,\g}},
  \label{eq:H_M_preview}
\end{align}
where $\c_{\mu s}\equiv\c_{0,\mu s}$ is a fermionic annihilation
operator for an atom in the motional ground state with nuclear spin
$\mu$ and orbital state $s$;
$\n_{\mu s}\equiv \c_{\mu s}^\dag \c_{\mu s}$ is a number operator;
and the sum is performed over all choices of nuclear spins $\mu_j$
with $j=1,2,\cdots,M$ for which all $\mu_j$ are distinct, or
equivalently all choices of $\mu_j$ for which the set $\set{\mu_j}$
contains $M$ elements, for a total of ${2I+1\choose M}\times\p{M!}$
nuclear spin combinations.  The coefficients $U_X$ can be expressed in
terms of the coupling constants of the bare two-body interaction
Hamiltonian (see Appendix \ref{sec:U_X}).  The key feature of the
$M$-body interactions in \eqref{eq:H_M_preview} is that they take a
form identical to the two-body interactions between motional
ground-state atoms, but with the addition of $M-2$ spectator atoms.
This form is a direct consequence of the SU($N$) symmetry of
underlying two-body interactions.

@@

As the bare two-body Hamiltonian $H_{\t{int}}$ is not diagonal with
respect to single-particle motional states, the problem of determining
many-body spectra and eigenstates nominally involves all motional
degrees of freedom of the atoms involved.  If we are interested in
addressing internal atomic states at zero temperature, however, in
principle we need only to consider the motional ground state
associated with each many-body internal atomic state.  We can then
ignore all excited motional states, as they will be neither thermally
occupied nor externally interrogated.  This procedure makes the
many-body problem more tractable by greatly reducing the
dimensionality of the relevant Hilbert space.

@@

two-body interaction.  On physical grounds, the net effective two-body
interaction must clearly be finite, but individual sums over excited
states in the loop diagrams of \eqref{eq:H_2_expansion} may generally
diverge \cite{johnson2012effective}.  These divergences ultimately
appear due to our use of effective field theory to describe
inter-atomic interactions in \eqref{eq:H_int_start}, rather than a
detailed microscopic description of two-atom scattering.  Divergences
of this sort are a generic feature of field theories, and can be dealt
with using standard techniques, in particular through renormalization.
We therefore renormalize our coupling constants by introducing

@@

via $G_X\to G_X+\tilde G_X$.  The values of counter-terms can be fixed
by enforcing that $G_X$ is equal to the net two-body interaction
strength, which has the added benefit of allowing us to express the
effective two-body interactions in terms of net, physically observable
two-body coupling constants on the right-hand side of
\eqref{eq:H_2_expansion}, rather than in terms of long sums at all
orders of the bare coupling constants.  Diagrammatically, this
renormalization condition is

@@

a vertex. To leading order, it follows that the counter-terms are
second order in the coupling constants, i.e.~$\tilde G\sim G^2$.  This
finding implies that counter-terms only come into play at
next-to-leading order (NLO) in the calculation of any effective
$M$-body interaction Hamiltonian.  By construction, these
counter-terms exactly cancel all terms beyond leading order in

@@

Second, the overview we provided in section \ref{sec:overview}, and
consequently our result in \eqref{eq:H_2}, neglects the effect of
momentum-dependent two-body scattering \cite{blume2002fermi}.  When
the effective range of momentum-dependent interactions is comparable
to the scattering lengths $a_X$, as is the case for ultracold
${}^{87}$Sr, these interactions are third order in the coupling
constants $G_X$.  Just as the $\O\p{G^2}$ counter-terms do not affect
$M$-body interactions until next-to-leading order (NLO), the
$\O\p{G^3}$ momentum-dependent interactions do not come into play
until next-next-leading order (NNLO).  Given that we develop our
low-energy effective theory through third order in the coupling
constants, these interactions will not appear in any of our three- and
four-body calculations, but they do need to be considered in the
calculation of pair-wise interaction energies.  The primary interest
of our work, however, concerns effective $M$-body interactions for
$M\ge3$; we therefore defer the calculation of momentum-dependent
two-body interactions to Appendix \ref{sec:momentum_dependence}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary of experimental procedures}
\label{sec:experiment_summary}
